Crowdsourcing and phylogenetic modelling reveal parrot tool use is not rare

### 2024-02-04 manuscript.tex
update manuscript and prepare for submission

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{tool use; technical intelligence; parrots; crowdsourcing; phylogenetic modelling\newline\indent Word count: 6548 words}
\usepackage{lineno}

@@

\abstract{%
Putatively rare behaviours, such as tool use, are challenging to study because absence of evidence can arise either from a species' inability to produce the behaviour or from insufficient research effort. Here, we tackle this challenge by combining crowdsourcing and phylogenetic modelling to estimate actual rates of tool use in parrots. Crowdsourcing revealed novel instances of tool use in 17 parrot species, more than doubling the confirmed number of tool-using parrot species from 11 (3\%) to 28 (7\%). Phylogenetic models incorporating these data together with measures of research effort and hypothesised predictors of tool use suggest that between 11\% and 17\% of parrot species may be tool users. These findings challenge the assumption that current research effort captures the distribution of putatively rare animal behaviours.
}

@@

\nolinenumbers

Our understanding of the evolution of animal behaviour is built on the
assumption that we have access to sufficient data (\emph{1}--\emph{3}). However, this is not always the case. Data on behaviours that are
rare, fleeting, or otherwise difficult to observe are likely to be patchy and
incomplete (\emph{4}, \emph{5}). Among species for which such behaviours
have not been observed, it can be difficult to differentiate between cases in
which the species is truly incapable of producing the behaviour and cases in
which the species is capable of producing the behaviour but the behaviour has
not yet been observed. Such a distinction can be critical for drawing
conclusions about the rarity and evolution of the behaviour in question.

Comparative work on the evolution of tool use is a paradigmatic example of this
issue. The initial discoveries of tool use in chimpanzees (\emph{6}),
birds (\emph{7}), dolphins (\emph{8}), and octopuses (\emph{9}) occurred
decades after significant advances on other more easily measurable aspects of
their biology. Since then, scholars have proposed a clear operational
definition of tool use applicable to all species --- the manipulation of an
unattached object as an extension of the animal's body to achieve a
goal (\emph{10}) --- and have used the distribution of species meeting this
definition to make various claims about the evolutionary drivers of tool use
behaviours. For example, based on the observation that bird species with
reported tool use tended to have larger brains, researchers have identified
higher relative brain size as a likely precondition for tool using
capabilities (\emph{11}--\emph{13}). These researchers
argue that larger brains are better able to integrate visual and somatosensory
information when innovating novel behaviours, such as tool use, in changing
environments (\emph{14}, \emph{15}). Similarly, researchers have used existing
reports of tool use in birds to debate the roles of generalist versus
specialist feeding strategies in driving the evolution of tool use, with some
arguing that feeding generalists require technical innovations to expand their
dietary niche (\emph{14}, \emph{16}, \emph{17}) and others arguing that
feeding specialists require technical innovations for extractive foraging of
specific foods (\emph{18}, \emph{19}).

However, before we can make claims of this kind, we need to know whether current
research effort in the literature is sufficient for robust conclusions to be
drawn about the evolution of tool use. In fact, evidence suggests that research
effort is often systematically biased towards particular taxonomic groups, parts
of the world that are easy to access, and species with life history traits that
make them easier to study, such as larger distribution ranges and population
sizes (\emph{20}). This is a crucial limitation because insufficient
observation may lead researchers to miss true instances of tool behaviours and
thus draw premature conclusions about the evolutionary drivers and origins of
tool use. While researchers have attempted to deal with this problem by
controlling for the number of scientific papers published on different species,
previous work has not yet attempted to quantify and explicitly model the
relationship between actual tool-using behaviour and what is reported in the
scientific literature. If more tool-using species exist than previously thought,
this could have important implications for theories of the evolutionary drivers
and origins of tool use and for our understanding of how rare this behaviour
actually is.

One potentially powerful method for quantifying actual rates of rare animal
behaviours is crowdsourcing (\emph{21}). In a crowdsourcing study, researchers
collect reports from the general public and/or collate and analyse videos posted
on social media platforms. This citizen science approach has been widely used in
ecology to monitor the distributional patterns of species (\emph{22}), but
has also recently been used to uncover a variety of rare animal behaviours,
including interspecies play in dogs (\emph{21}), novel problem-solving
behaviours in horses (\emph{23}), and socially-learned foraging innovations
in cockatoos (\emph{24}, \emph{25}). By casting the net wider than the
scientific literature, the crowdsourcing method can provide an indication of the
tool-using species that the literature might be missing.

Even after using this crowdsourcing approach, some tool-users could \emph{still}
remain unobserved. One principled framework for identifying these unobserved species
is to specify a causal model of the process that generates the observed data. We
propose one such causal model in Figure \ref{fig:plotDAG}. In this model, we
assume that the presence of tool use in the scientific literature (or in
crowdsourced reports) is caused by both unobserved tool use capabilities and the
number of published studies (or the number of crowdsourced reports) for any
given species. Tool users are more likely to be observed if they are well
studied, but understudied tool users may go undetected. Furthermore, based on
existing theories of the evolution of tool use (\emph{11}--\emph{19}), we propose that the
unobserved tool use capabilities are additionally caused by relative brain size,
feeding strategy, and shared phylogenetic ancestry. Expressing this causal model
as a statistical model can suggest further species which are likely to be
unobserved tool-users and, simultaneously, test existing theories of the
evolutionary drivers of tool use without incorrectly assuming that absence of
evidence is evidence of absence.

@@

Here, we apply these crowdsourcing and phylogenetic modelling approaches to tool
use in the parrot order. We focus on tool use in parrots for a number of
reasons. First, the scientific literature suggests that only a small proportion
of extant parrot species (11 out of 398; 3\%) use tools
(\emph{11}, \emph{26}--\emph{38}). Parrot tool use
thus provides an ideal test case for examining how robust sampling is in the
scientific literature. Second, parrots are highly popular as pets. Over 70\% of
all extant parrot species are bred in the aviculture industry and kept as pets
worldwide (\emph{39}--\emph{45}), enabling us to leverage the power of crowdsourcing on a social
media platform to search for evidence of tool use (\emph{21}). Third, detailed
data on relative brain sizes (\emph{46}--\emph{51}), feeding strategies (\emph{52}), and
shared ancestry (\emph{53}) exist for parrots, allowing us to fit the
statistical model implied by Figure \ref{fig:plotDAG} to the entire parrot
order.

@@

We surveyed the social media platform YouTube for video evidence of tool use in
parrots (see Methods for detailed search criteria). In our search, we used the
standard criteria for identifying tool use in the literature, defining ``true''
tool use behaviour as the manipulation of an unattached object as an extension
of the animal's body to achieve a goal (\emph{10}), while ``borderline'' tool
use involved the use of an object that was still attached to a substrate
(\emph{54}).

@@

According to YouTube video descriptions and owner comments,
45
of the individuals performing true tool use were males and
18
were females. No sex information was provided for the remaining
26
individuals. As owners provided no information on whether sex had been
established through genetic testing, and sexual dimorphism in parrots is
rare (\emph{55}, \emph{56}), we could not typically ascertain if descriptions
were accurate. It is unclear if the disproportionately large number of males in
the sample is a consequence of owners more readily assuming their parrots are
male when they have not been genetically tested, owners being more likely to own
or film male parrots, or male parrots exhibiting more true tool use behaviours
than female parrots.

@@

Survival cure models (\emph{57}), also known as split population
models (\emph{58}), are used to analyse the time to some event of interest
with the added assumption that a certain proportion of the population will never
experience the event, no matter how long they are measured for. These models
have been used to analyse a variety of right-censored outcomes, from cancer
relapse (\emph{57}) to criminal recidivism (\emph{58}). The data are
right-censored because some individuals will have experienced the event when
they are measured (e.g., disease onset, return to prison) while others will have
not experienced the event. For those who have not experienced the event, this
may be because (\emph{i}) the event has not happened to them yet or (\emph{ii}) the event
will \emph{never} happen to them. Survival cure models treat these two processes
separately.

@@

In our model, we infer the tool-using status of each species by allowing each
species to have their own probability of being a non-tool-user. Following our
causal model (Figure \ref{fig:plotDAG}), we predict these probabilities based
on feeding strategy, encephalisation quotient, and phylogenetic history (see
Methods for full model). The model additionally takes research effort into
account by allowing that, among species for which tool use is unobserved, all
else being equal those with fewer published papers and fewer video search hits
have a higher probability of being undetected tool users (fig.~
\ref{fig:plotSurvCure3}).

We found that this phylogenetic survival cure model was able to adequately
distinguish between species with and without evidence for tool use, with an
area-under-the-curve classification statistic of
0.95 (fig.\\
\ref{fig:plotSurvCure9}). To further estimate the accuracy of the model's
predictions, we also used a leave-one-species-out approach with known tool
users. For each of the 25 tool-using species that were represented on the
phylogeny and for which we had brain size and genomic data (we lacked data for
three tool-using species), we fitted the model to a modified dataset which set
tool use to be absent for the target species in both the scientific literature
and the video survey. Across 25 cross-validation models,
18 models (72\%)
continued to predict the target species as having a median posterior probability
of tool use that was within the range of all other tool users. This
classification rate was greater than the baseline classification rate of
26\%
for species without evidence of tool use in the full model
(38
of 149 species without
evidence of tool use had a median posterior probability of tool use that was
within the range of the tool-using species). Together, the area-under-the-curve
statistic and the leave-one-species-out approach suggest that the model is able
to adequately classify known tool users, with some error.

@@

The predicted probabilities from our phylogenetic survival cure model have
implications for inferences about the evolutionary drivers and origins of tool
use in the parrot order. Regarding the drivers of tool use hypothesised in
Figure \ref{fig:plotDAG}, the phylogenetic survival cure model revealed that
encephalisation quotient strongly positively predicted the probability of tool
use (median posterior log odds slope = 1.12, 95\%
CI {[}0.39
2.00{]}; total \emph{n} = 174
species; Figure \ref{fig:plotSurvCure4}). This helps explain the ranking in
Figure \ref{fig:plotSurvCure1}: the blue-eyed cockatoo has the largest relative
brain size in the dataset. We also found that feeding generalist species were
slightly more likely to be tool users, though the posterior difference between
generalists and specialists was quite uncertain (median posterior log odds
difference = 0.33, 95\% CI {[}
-1.13 1.76{]}; total
\emph{n} = 174 species). These results from the survival cure model
differed from the results of models fitted to pre-video-survey and
post-video-survey data without the survival cure component, which found no
effect of relative brain size and no difference between feeding strategies,
respectively (fig.~\ref{fig:plotComparison}).

@@

Regarding the origins of tool use, we fitted exploratory ancestral state
reconstruction models to the pre-video-survey data, the post-video-survey data,
and the predicted probabilities from the phylogenetic survival cure model. The
discoveries from our video survey and from our phylogenetic modelling increased
the likelihood that tool use was present in the most recent common ancestor for
several parrot genera. These include the amazon parrots native to the Americas
(\emph{Amazona}), the true white cockatoos and corellas found in South East Asia and
Australasia (\emph{Cacatua}), the kea and the kākā from New Zealand (\emph{Nestor}), and
the \emph{Poicephalus} genus native to Africa (Table \ref{tab:tableASR};
fig.~s \ref{fig:plotASR1} - \ref{fig:plotASR3}). These findings
suggest that species from each of these genera may share their tool use
capabilities via common descent from their respective common ancestors, rather
than via independent evolution within each genus or behavioural innovation
within species.

@@

Since the earliest anecdotes of parrots using tools by Wallace in the 1880s
(\emph{59}) and more systematic anecdotal reports in the 1970s (\emph{60}),
only 11 parrot species (3\% of all extant parrots) have been documented as tool
users in the scientific literature. Our study used crowdsourcing to identify 17
additional tool-using parrot species that are new to science, more than doubling
the overall count to 28 species (7\%). These species consisted of cockatoos
(\emph{Cacatuidae}), Old World parrots (\emph{Psittacinae}), and neotropical parrots
(\emph{Arinae}).

While parrots in the wild use tools to achieve a variety of
goals (\emph{34}, \emph{38}), the 17 additional species identified by our
video survey were exclusively observed using tools for self-scratching. This is
likely because preening is the only context in which captive parrots regularly
require tools, unlike wild parrots who may also require tools for extractive
foraging (\emph{34}, \emph{38}). Despite this difference in the type of tool
use observed, self-scratching is nonetheless a major category of animal tool
use and clearly meets the established criteria for tool use in the literature:
the manipulation of an unattached object as an extension of the animal's body to
achieve a goal (\emph{10}). It also meets the definition for a more complex
type of embodied tool use known as `tooling', the deliberate generation of a
mechanical interface by using an object to manipulate another target or
surface (\emph{61}). The fact that we found strong phylogenetic clustering
of self-scratching tool use from the video survey with other examples of tool
use from the literature supports a common underlying cognitive mechanism. In
line with this, some of the species of parrots that use tools for
self-scratching in captivity also use tools for other purposes in the wild
(\emph{30}, \emph{31}, \emph{34}, \emph{38}).

There was little evidence to suggest that the observations of self-scratching
tool use in the video survey were merely unintentional accidents or explicitly
trained behaviours. Individual parrots often used tools slowly and repetitively
over long periods of time, even across multiple different videos, suggesting
that their behaviour was not random or accidental (\emph{62}). Parrots
preferentially employed self-scratching tools on areas of their body that were
otherwise inaccessible, with 96\% of all instances involving scratching of the
head or neck, suggesting intentional tool use. For 60\% of the species in the
video survey, we found two or more videos of repetitive and sustained scratching
by different individuals of the same species in separate households, suggesting
that the manipulations were intentional and recurring events that did not
represent unusual stereotypies of any single individual. Regarding the
possibility of training or cueing from owners, over half of the videos contained
no evidence of human interaction aside from filming the behaviour. Humans only
handed parrots their tools in two of the videos, and none of the videos featured
owners directly rewarding tool use behaviours with food. Finally, the high
levels of phylogenetic signal in our data provides strong evidence that the
observations from our video survey reflect biologically-endowed capacities for
tool use rather than accidental or trained behaviours, which would likely appear
uniformly across the phylogeny.

Beyond the video survey, the strong phylogenetic signal in our dataset allowed
us to use phylogenetic information, along with other variables, to infer the
unobserved probabilities of tool use across the parrot order. Our phylogenetic
survival cure model incorporated information on phylogenetic history, research
effort, relative brain size, and feeding specialisation to rank parrot species
that were most likely to be undetected tool users. The sum of probabilities from
this model implied that between 15 and 41 of the species without observed
evidence of tool use are likely to be undetected tool users, suggesting that the
true proportion of tool users may be as high as 17\%.

@@

Second, in terms of the evolution of tool use in parrots, our study challenges a
key assumption made in the literature to date: that only a minority of parrots
are tool users (\emph{30}, \emph{31}, \emph{34}, \emph{37}, \emph{38}).
The paucity of evidence for tool use across parrots in the literature initially
implied that the capacity for tool use may have evolved independently in
different parrot species. Our discovery of the widespread distribution of tool
use across the parrot phylogeny, along with the strong phylogenetic signal in
this expanded dataset, challenges this and suggests that, at least for some
parrot clades, the capacity for tool use might be a homologous trait that has
been evolutionarily conserved. Our exploratory ancestral state reconstruction
analysis provides preliminary support for this hypothesis, revealing an
increased probability of tool use among the most recent common ancestors for the
\emph{Amazona}, \emph{Cacatua}, \emph{Nestor}, and \emph{Poicephalus} genera. Even at this
preliminary stage, our analysis therefore raises an alternative hypothesis for
the observed tool use in \emph{Cacatua} (\emph{30}, \emph{38}) and
\emph{Nestor} (\emph{26}--\emph{29}, \emph{31}, \emph{34}), namely that tool behaviours have arisen due to the
common ancestor having the capacity to use tools, rather than from independent
evolution or behavioural innovation within species.

Third, our results support existing theories of the drivers of tool use. We
found that encephalisation was strongly positively related to the probability of
tool use in our phylogenetic model, supporting previous theories linking
relative brain size to increased tool innovation in birds (\emph{11}--\emph{13}) and primates (\emph{63}). We acknowledge that
encephalisation quotient is not a perfect measure due to measurement error and
challenges with interpretation (\emph{3}, \emph{4}). However, encephalisation
quotient has much greater coverage across the parrot phylogeny than more
fine-grained measures like whole neuron count (\emph{64}) and there is no reason
to think that measurement error would produce the consistent patterns across our
study and prior work. To understand the causal mechanisms responsible for these
relationships, we encourage further work on the specific neural correlates of
technical intelligence in parrots, e.g., (\emph{65}). In our phylogenetic
model, we also found that tool use was somewhat more likely among feeding
generalists compared to feeding specialists, although this difference was
uncertain. This trend supports previous suggestions that increased cognitive
abilities and technical innovation rates are required to expand a generalist
species' dietary niche (\emph{16}, \emph{18}, \emph{66}). However, the trend contradicts theories linking tool use to
dietary specialisation, whereby species eating specific foods that require
extractive foraging have higher cognitive ability and are especially prone to
using tools (\emph{18}).

While missing data imputation is becoming more common in phylogenetic
analyses (\emph{67}), the important distinction between absence of
evidence and evidence of absence has not been given as much attention. Our
phylogenetic analysis provides one approach to this problem by distinguishing
between true absences of tool use and absences of tool use due to a lack of
research effort in the scientific literature or in crowdsourced videos. To
achieve this, we explicitly modelled the measurement of the outcome variable
along a research effort time series, such that species with lower research
effort in the literature or in videos were likely to be censored. In line with
our causal model, we also included relative brain size, feeding strategy, and
phylogenetic history as predictors of unobserved tool use. We encourage
researchers to test this model by directing future study efforts towards the
parrot species with the highest probabilities of being undetected tool-users.
Future work should also refine the causal model in Figure \ref{fig:plotDAG} to
provide more certain estimates of tool use probabilities, either by including
additional predictor variables or modelling further causes of measurement error
in the taxonomic record, such as species abundance and geographic
accessibility (\emph{20}).

In conclusion, we have shown that the scientific literature has insufficiently
captured the full distribution of tool use in the parrot order. Our
crowdsourcing survey more than doubled the number of known tool using parrot
species from 11 to 28, and our phylogenetic model suggested that the true
proportion of parrot tool users could be as high as 17\% of all species in this
order. These discoveries have implications for theories of the evolutionary
drivers and origins of tool use in parrots. Beyond parrot tool use, the
crowdsourcing and phylogenetic methods used in this study have the potential to
be applied to other rarely observed behaviours, including tool use in other
taxa (\emph{68}), rhythmic entrainment in birds (\emph{69}--\emph{72}), teasing behaviours in primates (\emph{73}), and
tactical deception across all animals (\emph{74}--\emph{76}).
We hope that these methods will continue to uncover a diverse array of ephemeral
behaviours that have as yet gone undetected in the scientific literature.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{tool use; technical intelligence; parrots; crowdsourcing; phylogenetic modelling}
\usepackage{lineno}

@@

\abstract{%
Putatively rare behaviours, such as tool use, are challenging to study because absence of evidence can arise either from a species' inability to produce the behaviour or from insufficient research effort. Here, we tackle this challenge by combining crowdsourcing and phylogenetic modelling to estimate actual rates of tool use in parrots. Crowdsourcing on a social media platform revealed novel instances of tool use in 17 parrot species, more than doubling the confirmed number of tool-using parrot species from 11 (3\%) to 28 (7\%). Phylogenetic modelling ranked additional species that are most likely to be unobserved tool users, suggesting that between 11\% and 17\% of extant parrot species may be tool users. These discoveries have implications for inferences about the evolutionary drivers and origins of tool use in parrots, revealing associations with relative brain size and feeding generalism and indicating several genera where tool use was likely an ancestral trait. Overall, our findings challenge the assumption that current sampling effort captures the full distribution of putatively rare animal behaviours. Combining our sampling and analysis methods offers a fruitful approach for investigating the distribution, drivers, and origins of other rare behaviours.
}

@@

\linenumbers

Our understanding of the evolution of animal behaviour is built on the
assumption that we have access to sufficient data (1--3). However, this is not always the case. Data on behaviours that are
rare, fleeting, or otherwise difficult to observe are likely to be patchy and
incomplete (4, 5). Among species for which such behaviours
have not been observed, it can be difficult to differentiate between cases in
which the species is truly incapable of producing the behaviour and cases in
which the species is capable of producing the behaviour but the behaviour has
not yet been observed. Such a distinction can be critical for drawing
conclusions about the rarity and evolution of the behaviour in question.

Comparative work on the evolution of tool use is a paradigmatic example of this
issue. The initial discoveries of tool use in chimpanzees (6),
birds (7), dolphins (8), and octopuses (9) occurred
decades after significant advances on other more easily measurable aspects of
their biology. Since then, scholars have proposed a clear operational
definition of tool use applicable to all species --- the manipulation of an
unattached object as an extension of the animal's body to achieve a
goal (10) --- and have used the distribution of species meeting this
definition to make various claims about the evolutionary drivers of tool use
behaviours. For example, based on the observation that bird species with
reported tool use tended to have larger brains, researchers have identified
higher relative brain size as a likely precondition for tool using
capabilities (11--13). These researchers
argue that larger brains are better able to integrate visual and somatosensory
information when innovating novel behaviours, such as tool use, in changing
environments (14, 15). Similarly, researchers have used existing
reports of tool use in birds to debate the roles of generalist versus
specialist feeding strategies in driving the evolution of tool use, with some
arguing that feeding generalists require technical innovations to expand their
dietary niche (14, 16, 17) and others arguing that
feeding specialists require technical innovations for extractive foraging of
specific foods (18, 19).

However, before we can make claims of this kind, we need to know whether current
research effort in the literature is sufficient for robust conclusions to be
drawn about the evolution of tool use. In fact, evidence suggests that research
effort is often systematically biased towards particular taxonomic groups, parts
of the world that are easy to access, and species with life history traits that
make them easier to study, such as larger distribution ranges and population
sizes (20). This is a crucial limitation because insufficient
observation may lead researchers to miss true instances of tool behaviours and
thus draw premature conclusions about the evolutionary drivers and origins of
tool use. While researchers have attempted to deal with this problem by
controlling for the number of scientific papers published on different species,
previous work has not yet attempted to quantify and explicitly model the
relationship between actual tool-using behaviour and what is reported in the
scientific literature. If more tool-using species exist than previously thought,
this could have important implications for theories of the evolutionary drivers
and origins of tool use and for our understanding of how rare this behaviour
actually is.

One potentially powerful method for quantifying actual rates of rare animal
behaviours is crowdsourcing (21). In a crowdsourcing study, researchers
collect reports from the general public and/or collate and analyse videos posted
on social media platforms. This citizen science approach has been widely used in
ecology to monitor the distributional patterns of species (22), but
has also recently been used to uncover a variety of rare animal behaviours,
including interspecies play in dogs (21), novel problem-solving
behaviours in horses (23), and socially-learned foraging innovations
in cockatoos (24, 25). By casting the net wider than the
scientific literature, the crowdsourcing method can provide an indication of the
tool-using species that the literature might be missing.

Even after using this crowdsourcing approach, some tool-users could \emph{still}
remain unobserved. One principled framework for identifying these unobserved species
is to specify a causal model of the process that generates the observed data. We
propose one such causal model in Figure \ref{fig:plotDAG}. In this model, we
assume that the presence of tool use in the scientific literature (or in
crowdsourced reports) is caused by both unobserved tool use capabilities and the
number of published studies (or the number of crowdsourced reports) for any
given species. Tool users are more likely to be observed if they are well
studied, but understudied tool users may go undetected. Furthermore, based on
existing theories of the evolution of tool use (11--19), we propose that the
unobserved tool use capabilities are additionally caused by relative brain size,
feeding strategy, and shared phylogenetic ancestry. Expressing this causal model
as a statistical model can suggest further species which are likely to be
unobserved tool-users and, simultaneously, test existing theories of the
evolutionary drivers of tool use without incorrectly assuming that absence of
evidence is evidence of absence.

@@

Here, we apply these crowdsourcing and phylogenetic modelling approaches to tool
use in the parrot order. We focus on tool use in parrots for a number of
reasons. First, the scientific literature suggests that only a small proportion
of extant parrot species (11 out of 398; 3\%) use tools
(11, 26--38). Parrot tool use
thus provides an ideal test case for examining how robust sampling is in the
scientific literature. Second, parrots are highly popular as pets. Over 70\% of
all extant parrot species are bred in the aviculture industry and kept as pets
worldwide (39--45), enabling us to leverage the power of crowdsourcing on a social
media platform to search for evidence of tool use (21). Third, detailed
data on relative brain sizes (46--51), feeding strategies (52), and
shared ancestry (53) exist for parrots, allowing us to fit the
statistical model implied by Figure \ref{fig:plotDAG} to the entire parrot
order.

@@

We surveyed the social media platform YouTube for video evidence of tool use in
parrots (see Methods for detailed search criteria). In our search, we used the
standard criteria for identifying tool use in the literature, defining ``true''
tool use behaviour as the manipulation of an unattached object as an extension
of the animal's body to achieve a goal (10), while ``borderline'' tool
use involved the use of an object that was still attached to a substrate
(54).

@@

According to YouTube video descriptions and owner comments,
45
of the individuals performing true tool use were males and
18
were females. No sex information was provided for the remaining
26
individuals. As owners provided no information on whether sex had been
established through genetic testing, and sexual dimorphism in parrots is
rare (55, 56), we could not typically ascertain if descriptions
were accurate. It is unclear if the disproportionately large number of males in
the sample is a consequence of owners more readily assuming their parrots are
male when they have not been genetically tested, owners being more likely to own
or film male parrots, or male parrots exhibiting more true tool use behaviours
than female parrots.

@@

Survival cure models (57), also known as split population
models (58), are used to analyse the time to some event of interest
with the added assumption that a certain proportion of the population will never
experience the event, no matter how long they are measured for. These models
have been used to analyse a variety of right-censored outcomes, from cancer
relapse (57) to criminal recidivism (58). The data are
right-censored because some individuals will have experienced the event when
they are measured (e.g., disease onset, return to prison) while others will have
not experienced the event. For those who have not experienced the event, this
may be because (\emph{i}) the event has not happened to them yet or (\emph{ii}) the event
will \emph{never} happen to them. Survival cure models treat these two processes
separately.

@@

In our model, we infer the tool-using status of each species by allowing each
species to have their own probability of being a non-tool-user. Following our
causal model (Figure \ref{fig:plotDAG}), we predict these probabilities based
on feeding strategy, encephalisation quotient, and phylogenetic history (see
Methods for full model). The model additionally takes research effort into
account by allowing that, among species for which tool use is unobserved, all
else being equal those with fewer published papers and fewer video search hits
have a higher probability of being undetected tool users (Supplementary Figure
\ref{fig:plotSurvCure3}).

We found that this phylogenetic survival cure model was able to adequately
distinguish between species with and without evidence for tool use, with an
area-under-the-curve classification statistic of
0.95 (Supplementary Figure
\ref{fig:plotSurvCure9}). To further estimate the accuracy of the model's
predictions, we also used a leave-one-species-out approach with known tool
users. For each of the 25 tool-using species that were represented on the
phylogeny and for which we had brain size and genomic data (we lacked data for
three tool-using species), we fitted the model to a modified dataset which set
tool use to be absent for the target species in both the scientific literature
and the video survey. Across 25 cross-validation models,
18 models (72\%)
continued to predict the target species as having a median posterior probability
of tool use that was within the range of all other tool users. This
classification rate was greater than the baseline classification rate of
26\%
for species without evidence of tool use in the full model
(38
of 149 species without
evidence of tool use had a median posterior probability of tool use that was
within the range of the tool-using species). Together, the area-under-the-curve
statistic and the leave-one-species-out approach suggest that the model is able
to adequately classify known tool users, with some error.

@@

The predicted probabilities from our phylogenetic survival cure model have
implications for inferences about the evolutionary drivers and origins of tool
use in the parrot order. Regarding the drivers of tool use hypothesised in
Figure \ref{fig:plotDAG}, the phylogenetic survival cure model revealed that
encephalisation quotient strongly positively predicted the probability of tool
use (median posterior log odds slope = 1.12, 95\%
CI {[}0.39
2.00{]}; total \emph{n} = 174
species; Figure \ref{fig:plotSurvCure4}). This helps explain the ranking in
Figure \ref{fig:plotSurvCure1}: the blue-eyed cockatoo has the largest relative
brain size in the dataset. We also found that feeding generalist species were
slightly more likely to be tool users, though the posterior difference between
generalists and specialists was quite uncertain (median posterior log odds
difference = 0.33, 95\% CI {[}
-1.13 1.76{]}; total
\emph{n} = 174 species). These results from the survival cure model
differed from the results of models fitted to pre-video-survey and
post-video-survey data without the survival cure component, which found no
effect of relative brain size and no difference between feeding strategies,
respectively (Supplementary Figure \ref{fig:plotComparison}).

@@

Regarding the origins of tool use, we fitted exploratory ancestral state
reconstruction models to the pre-video-survey data, the post-video-survey data,
and the predicted probabilities from the phylogenetic survival cure model. The
discoveries from our video survey and from our phylogenetic modelling increased
the likelihood that tool use was present in the most recent common ancestor for
several parrot genera. These include the amazon parrots native to the Americas
(\emph{Amazona}), the true white cockatoos and corellas found in South East Asia and
Australasia (\emph{Cacatua}), the kea and the kākā from New Zealand (\emph{Nestor}), and
the \emph{Poicephalus} genus native to Africa (Table \ref{tab:tableASR};
Supplementary Figures \ref{fig:plotASR1} - \ref{fig:plotASR3}). These findings
suggest that species from each of these genera may share their tool use
capabilities via common descent from their respective common ancestors, rather
than via independent evolution within each genus or behavioural innovation
within species.

@@

Since the earliest anecdotes of parrots using tools by Wallace in the 1880s
(59) and more systematic anecdotal reports in the 1970s (60),
only 11 parrot species (3\% of all extant parrots) have been documented as tool
users in the scientific literature. Our study used crowdsourcing to identify 17
additional tool-using parrot species that are new to science, more than doubling
the overall count to 28 species (7\%). These species consisted of cockatoos
(\emph{Cacatuidae}), Old World parrots (\emph{Psittacinae}), and neotropical parrots
(\emph{Arinae}). Beyond the crowdsourcing survey, the strong phylogenetic signal in
our dataset allowed us to use phylogenetic information, along with other
variables, to infer the unobserved probabilities of tool use across the parrot
order. Our phylogenetic survival cure model incorporated information on
phylogenetic history, research effort, relative brain size, and feeding
specialisation to rank parrot species that were most likely to be undetected
tool users. The sum of probabilities from this model implied that between 15
and 41 of the species without observed evidence of tool use are likely to be
undetected tool users, suggesting that the true proportion of tool users may be
as high as 17\%.

@@

Second, in terms of the evolution of tool use in parrots, our study challenges a
key assumption made in the literature to date: that only a minority of parrots
are tool users (30, 31, 34, 37, 38).
The paucity of evidence for tool use across parrots in the literature initially
implied that the capacity for tool use may have evolved independently in
different parrot species. Our discovery of the widespread distribution of tool
use across the parrot phylogeny, along with the strong phylogenetic signal in
this expanded dataset, challenges this and suggests that, at least for some
parrot clades, the capacity for tool use might be a homologous trait that has
been evolutionarily conserved. Our exploratory ancestral state reconstruction
analysis provides preliminary support for this hypothesis, revealing an
increased probability of tool use among the most recent common ancestors for the
\emph{Amazona}, \emph{Cacatua}, \emph{Nestor}, and \emph{Poicephalus} genera. Even at this
preliminary stage, our analysis therefore raises an alternative hypothesis for
the observed tool use in \emph{Cacatua} (30, 38) and
\emph{Nestor} (26--29, 31, 34), namely that tool behaviours have arisen due to the
common ancestor having the capacity to use tools, rather than from independent
evolution or behavioural innovation within species.

Third, our results support existing theories of the drivers of tool use. We
found that encephalisation was strongly positively related to the probability of
tool use in our phylogenetic model, supporting previous theories linking
relative brain size to increased tool innovation in birds (11--13) and primates (61). We acknowledge that
encephalisation quotient is not a perfect measure due to measurement error and
challenges with interpretation (3, 4). However, encephalisation
quotient has much greater coverage across the parrot phylogeny than more
fine-grained measures like whole neuron count (62) and there is no reason
to think that measurement error would produce the consistent patterns across our
study and prior work. To understand the causal mechanisms responsible for these
relationships, we encourage further work on the specific neural correlates of
technical intelligence in parrots, e.g., (63). In our phylogenetic
model, we also found that tool use was somewhat more likely among feeding
generalists compared to feeding specialists, although this difference was
uncertain. This trend supports previous suggestions that increased cognitive
abilities and technical innovation rates are required to expand a generalist
species' dietary niche (16, 18, 64). However, the trend contradicts theories linking tool use to
dietary specialisation, whereby species eating specific foods that require
extractive foraging have higher cognitive ability and are especially prone to
using tools (18).

All instances of tool use in our crowdsourcing survey met the established
criteria for tool use in the literature: the manipulation of an unattached
object as an extension of the animal's body to achieve a goal (10).
However, a striking feature of the dataset is that the 17 additional species
identified in the video survey were exclusively observed using tools for
self-scratching. While parrots in the wild use tools to achieve a variety of
goals, including extractive foraging and preening (34, 38),
parrots in captivity only require tools to achieve the latter goal. Despite this
difference in the type of tool use observed, self-scratching is nonetheless a
major category of animal tool use. Self-scratching also meets the definition for
a more complex type of embodied tool use known as `tooling', the deliberate
generation of a mechanical interface by using an object to manipulate another
target or surface (65). This makes it difficult to dismiss
self-scratching as unimportant: to do so would require a redefinition of
tool-use in non-human animals. Moreover, the fact that we found strong
phylogenetic clustering of self-scratching tool use from the video survey with
other examples of tool use from the literature supports a common underlying
cognitive mechanism. In line with this, some of the species of parrots that use
tools for self-scratching in captivity also use tools for other purposes in the
wild (30, 31, 34, 38).

There was little evidence to suggest that the observations of self-scratching
tool use in the video survey were merely unintentional accidents or explicitly
trained behaviours. Individual parrots often used tools slowly and repetitively
over long periods of time, even across multiple different videos, suggesting
that their behaviour was not random or accidental (66). Parrots
preferentially employed self-scratching tools on areas of their body that were
otherwise inaccessible, with 96\% of all instances involving scratching of the
head or neck, suggesting intentional tool use. For 60\% of the species in the
video survey, we found two or more videos of repetitive and sustained scratching
by different individuals of the same species in separate households, suggesting
that the manipulations were intentional and recurring events that did not
represent unusual stereotypies of any single individual. Regarding the
possibility of training or cueing from owners, over half of the videos contained
no evidence of human interaction aside from filming the behaviour. Humans only
handed parrots their tools in two of the videos, and none of the videos featured
owners directly rewarding tool use behaviours with food. Finally, the high
levels of phylogenetic signal in our data provide strong evidence that the
observations from our video survey reflect biologically-endowed capacities for
tool use rather than accidental or trained behaviours, which would likely appear
uniformly across the phylogeny.
